---
title: "HPI Bayesian Course"
output: 
  html_notebook:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float: true
---

# How to follow

All code should be added into the ***Terminal*** tab (in the Console pane in RStudio). Alternatively, you don't need to use the Terminal at all, as RStudio adds a Git tab to the Environment pane. For a much more in-depth (and better) guide, follow the amazing https://happygitwithr.com/index.html by the even more amazing Jenny Bryan.

# Taken for granted

I take for granted:

- git is already installed on my machine
- I have a GitHub account
- already got RStudio set up with an SSH key (Tools > Global Options > Git). 
  - with an SSH key, I should always make sure to copy the SSH url and not the HTTPS (which would require a Personal Access Token (PAT))! 

If these steps haven't already been taken, check out the first chapter from https://happygitwithr.com/index.html

If you've done all that, follow along from here (following https://happygitwithr.com/rstudio-git-github.html#rstudio-git-github)

# Setting up

Steps:

1. Create a repo in GitHub
  - make sure to click 'create README.md'
2. Create a new RProject with Version Control
  - add the URL from your GitHub repo, press the green 'Clone <>' button
  - copy the url, but  make sure you choose the relevant format! SSH if you've got an SSH key set up, or HTTPS if you're using a personal access token (PAT)
3. stage (add), commit, some arbitray changes
  - using either the Git tab, or preferably using the Terminal (see below for examples)

```{r, eval = F, echo = T}
# add a line to your README
echo "This is a line from RStudio" >> README.md
# check the status
git status
# stage the change
git add README.md
# commit the change and add a message
git commit -m "first commit from RStudio"
# push the change to GitHub
git push
# now, go refresh your GH repo and see the change
```

## Other commands

```{r, eval = F, echo = T}
# print working directory; should be to your project folder
pwd
# change directory, in case not your project folder
cd
# list all files in WD
ls
# list all files including ignord ones
ls -a
# check status
status
# add file/folder to gitignore (so git will ignore it)
echo Slides/ >> .gitignore
```

# Keeping up changes

If you've already started working on the project locally, you can just drag-and-drop your files/folders into the Terminal

```{r, eval = F, echo = T}
# stage change (need to do it every time)
add filename.Rmd
# commit these changes to git
commit -m "message about commit"
# check the status
status
# push these changes to remote repo
push
```

Alternatively, you don't need to use the Terminal at all (but it's good practice to know how to in order to understand the logic/workflow). In the 'Environment' pane, you should see a 'Git' tab (if git is indeed set up). You can tick the 'Staged' button for all files whose changes you want to commit. Then click 'commit'. A message window will come up, add your message and click 'Commit'. Then, Click 'Push' to push to GH.

