---
title: "Graded homeworks"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

# Set options {-}

```{r setup, include=FALSE}
# set global knit options
knitr::opts_chunk$set(echo = T,
                      eval = T,
                      error = F,
                      warning = F,
                      message = F)

# suppress scientific notation
options(scipen=999)

# list of required packages
packages <- c( #"SIN", # this package was removed from the CRAN repository
               "MASS", "dplyr", "tidyr", "purrr", "extraDistr", "ggplot2", "loo", "bridgesampling", "brms", "bayesplot", "tictoc", "hypr", "bcogsci", "papaja", "grid", "kableExtra", "gridExtra", "lme4", "cowplot", "pdftools", "cmdstanr", "rootSolve", "rstan"
  )

# NB: if you haven't already installed bcogsci through devtools, it won't be loaded
## Now load or install & load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# this is also required, taken from the textbook

## Save compiled models:
rstan_options(auto_write = FALSE)
## Parallelize the chains using all the cores:
options(mc.cores = parallel::detectCores())
# To solve some conflicts between packages
select <- dplyr::select
extract <- rstan::extract
```

# Week 1

#### Question 1 (4.0 Pts)  {-}
What are the constraints on probability? Choose the correct answers:

```{r, eval = F}
If two events A and B are mutually exclusive, the events are called disjoint events. The probability of two disjoint events A or B happening is: p(A or B) = p(A) + p(B).

## A ***or*** B! but we want to know 
```


- The probability that several mutually exclusive events occur is the sum of the probabilities of each of these events.
- Several events can have the same probability. ***YES***
- The probabilities of all possible events in the entire sample space sum up to 1. ***YES***
- An event can have more than one probability.
- The probability of an event must lie between 0 and 1, including 0 and 1. ***YES***
- Several events in one sample space can have a probability of 1.
- The probability of an event A given another event B is always equal to the probability of the event A occurring.
- All events in the sample space are equally likely.

#### Question 2  {-}
In a random variable X that has as a PDF the standard normal distribution, what is the probability of observing the value x=1?

```{r}
# no code needed, this is a trick! with a continuous RV the chances of observing an exact value is always 0!
```

#### Question 3 (2.0 Pts)  {-}  {-}
Consider a discrete bivariate distribution. Is it generally true that P(X=a | Y=b) = P(Y=b | X=a)

#### Question 4 (1.0 Pts)  {-}
In a standard normal distribution with mean 0 and standard deviation 1, the probability of obtaining a value greater than 2 is:

```{r}
pnorm(2,
      lower.tail=F)
pnorm(-2)
# the same as obtaining a value less than -2
```

#### Question 5 (1.0 Pts)  {-}
Given a standard normal distribution, what is the probability of getting a value higher than 2.5?

```{r}
round(
  pnorm(q=2.5,
        lower.tail=F),
  3)
```

#### Question 6 (1.0 Pts)  {-}
Consider participating in a lottery ten times and winning 5 times. What is the maximum likelihood estimate of the probability of winning?

```{r}
# MLE = theta-hat = k/n = 
  5/10
```


#### Question 7 (1.0 Pts)  {-}
Is it true that the maximum likelihood estimate (MLE) from a particular sample of data always gives us an accurate estimate of the parameter of interest?

#### Question 8 (2.0 Pts)  {-}
Consider participating in a lottery 100 times and winning 50 times. The number of wins out of 100 attempts is the random variable of interest. Estimate the variance of this random variable.

```{r}
100 * .5 * (1-.5)
```

#### Question 9 (2.0 Pts)  {-}
Consider two independent random variables: 
X ~ N(mu_x, sigma_x) and Y ~ N(mu_y, sigma_y).

How is the bivariate distribution of X and Y specified?