---
title: "Course quizzes"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

# Set options {-}

```{r setup, include=FALSE}
# set global knit options
knitr::opts_chunk$set(echo = T,
                      eval = T,
                      error = F,
                      warning = F,
                      message = F)

# suppress scientific notation
options(scipen=999)

# list of required packages
packages <- c( #"SIN", # this package was removed from the CRAN repository
               "MASS", "dplyr", "tidyr", "purrr", "extraDistr", "ggplot2", "loo", "bridgesampling", "brms", "bayesplot", "tictoc", "hypr", "bcogsci", "papaja", "grid", "kableExtra", "gridExtra", "lme4", "cowplot", "pdftools", "cmdstanr", "rootSolve", "rstan"
  )

# NB: if you haven't already installed bcogsci through devtools, it won't be loaded
## Now load or install & load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# this is also required, taken from the textbook

## Save compiled models:
rstan_options(auto_write = FALSE)
## Parallelize the chains using all the cores:
options(mc.cores = parallel::detectCores())
# To solve some conflicts between packages
select <- dplyr::select
extract <- rstan::extract
```

# Week 1

## Function summary notes {-}

d-p-q-r family of functions

+-----------+-------------------------+--------------------------------------------------+
| RV        | Function(arguments)     | Outcome                                          |
+===========+=========================+==================================================+
| Bernoulli | `rbern(n,prob)`         | generate random data                             |
|           +-------------------------+--------------------------------------------------+
|           | `dbern(x,prob)`         | PMF: ***probability*** of outcome x              |
|           +-------------------------+--------------------------------------------------+
|           | `pbern(q,prob)`         | CMF: cumulative PMF of <=q                       |
+-----------+-------------------------+--------------------------------------------------+
| Binomial  | `rbinom(n, size, prob)` | generate random data with n = of successes       |
|           +-------------------------+--------------------------------------------------+
|           | `dbinom(q,size,prob)`   | PMF: ***probability*** of outcome x              |
|           +-------------------------+--------------------------------------------------+
|           | `pbinom(q,size,prob)`   | CMF: cumulative PMF of <=q                       |
|           +-------------------------+--------------------------------------------------+
|           | `qbinom(p,prob,size)`   | inverse CMF: cumulative PMF of >=x               |
+-----------+-------------------------+--------------------------------------------------+
| normal    | `rnorm(n,mean,sd)`      | generate random data                             |
|           +-------------------------+--------------------------------------------------+
|           | `dnorm(x,mean,sd)`      | PDF: ***density*** of outcome x                  |
|           +-------------------------+--------------------------------------------------+
|           | `pnorm(q,mean,sd)`      | CDF: cumulative PDF of <=q                       |
|           +-------------------------+--------------------------------------------------+
|           | `pnorm(q,mean,sd,`      | inverse CDF: cumulative PDF of >=q               |
|           |       `lower.tail=F)`   |                                                  |
|           +-------------------------+--------------------------------------------------+
|           | `qnorm(p,prob,size)`    | Compute quantiles corresponding to probabilities |
+-----------+-------------------------+--------------------------------------------------+

## Quiz: Binomial distribution

Q1: Consider participating in a lottery ten times. Each time the probability of winning a prize is 0.10. What is the probability of winning exactly 5 times?

```{r}
# use dbinom to compute PMF
round(
  dbinom(5, # produce probs for 5 successes
       size = 10, # out of 10 tries/trials
       prob = .1 # with a prob of success on each trial = .1
       ),
3)
```

Q2: Consider lending 10 books from a library. The probability of getting a damaged book is 0.15. Compute the cumulative probability of having 2 or fewer damaged books rounded to three digits.

```{r}
# use pbinom to compute CMF
round(
  pbinom(2, # successes
    size = 10, # out of 10
    prob = .15 # with prob of .15
    
  ),
3)
```

## Quiz: Normal distribution

Q1: Given a standard normal distribution, what is the probability of getting a value lower than -3?

```{r}
pnorm(-3)
```

Q2: Given a standard normal distribution, what is the probability of getting a value higher than -3?

```{r}
pnorm(-3, lower.tail=F)
```

## Quiz: Bivariate distribution
  
```{r}
pxy41 <- .096
py1 <- sum(.031,.053,.086,.096,.147,.153,.142)

px4y1 <- pxy41/py1; round(px4y1,3)
```

And the exercise from Slide 9:

```{r}
# first compute the probabilities of y=0 and y=1
py0 <- sum(.018,.023,.04,.043,.063,.049,.055)
py1 <- sum(.031,.053,.086,.096,.147,.153,.142)

## now compute probabilities of x1:7 given y = 0
px1y0 <- .018/py0
px2y0 <- .023/py0
px3y0 <- .04/py0
px4y0 <- .043/py0
px5y0 <- .063/py0
px6y0 <- .049/py0
px7y0 <- .055/py0
## sum should equal 1
pxy0 <- sum(px1y0,px2y0,px3y0,px4y0,px5y0,px6y0,px7y0)

## now compute probabilities of x1:7 given y = 1
px1y1 <- .031/py1
px2y1 <- .053/py1
px3y1 <- .086/py1
px4y1 <- .096/py1
px5y1 <- .147/py1
px6y1 <- .153/py1
px7y1 <- .142/py1
# sum should equal 1
pxy1 <- sum(px1y1,px2y1,px3y1,px4y1,px5y1,px6y1,px7y1)

px17y0 <- c(px1y0,px2y0,px3y0,px4y0,px5y0,px6y0,px7y0, pxy0)
px17y1 <- c(px1y1,px2y1,px3y1,px4y1,px5y1,px6y1,px7y1, pxy1)

# create data frame for table
px17y10 <- as.data.frame(rbind(px17y0,px17y1))
```

```{r, results = "asis"}
# print table
knitr::kable(px17y10,
             digits = 3,
             col.names =
              c("x=1","x=2","x=3","x=4",
                "x=5","x=6","x=7","sum"),
             # row.names = 
             #   c("P(x=1:7|y=0)", "P(x=1:7|y=1)")
             ) %>%
  kable_styling()

```

## Self-test: week 1

### Question 1 (1.0 Pts) {-}
Compute the probability of obtaining a value less than or equal to 5 in a normal distribution with mean 3 and standard deviation 1.

```{r}
round(pnorm(q = 5,
      mean = 3,
      sd = 1),
      3)
```

###  Question 2 (1.0 Pts) {-}
Compute the probability of obtaining a value less than or equal to 5 in a normal distribution with mean 3 and standard deviation 3.

```{r}
round(pnorm(q = 5,
      mean = 3,
      sd = 3),
      3)
```

### Question 3 (1.0 Pts) {-}
Compute the probability of obtaining a value more than 5 in a normal distribution with mean 3 and standard deviation 1.

```{r}
round(pnorm(q = 5,
      mean = 3,
      sd = 1,
      lower.tail = F),
      3)
```

### Question 4 (1.0 Pts) {-}
Compute the probability of obtaining a value more than or equal to 5 in a normal distribution with mean 4 and standard deviation 3.

```{r}
round(pnorm(q = 5,
      mean = 4,
      sd = 3,
      lower.tail = F),
      3)
```

### Question 5 (1.0 Pts) {-}
Compute the probability of obtaining a value between 2 and 5 in a normal distribution with mean 3 and standard deviation 1.

```{r}
round(pnorm(5,3,1)-pnorm(2,3,1),
      3)
```

### Question 6 (1.0 Pts) {-}
Compute the probability of obtaining a value between 1 and 1.5 in a normal distribution with mean 3 and standard deviation 1.

```{r}
round(pnorm(1.5,3,1) - 
        pnorm(1,3,1),
      3)
```

### Question 7 (1.0 Pts) {-}
Compute the probability of obtaining a value between 1 and 1.5 in a standard normal distribution.

```{r}
round(pnorm(1.5,0,1) - 
        pnorm(1,0,1),
      3)
```

### Question 8 (1.0 Pts) {-}
Compute the probability of obtaining 4 or fewer successes in 15 trials following a binomial distribution with probability of success being 0.5.

```{r}
round(
  pbinom(4,15,.5),
  3)
```

### Question 9 (1.0 Pts) {-}
Compute the probability of obtaining 3, 4, or 5 successes in 15 trials following a binomial distribution with probability of success being 0.5.

```{r}
round(
  sum(
    dbinom(3:5,15,.5)
    ),
  3)
```

### Question 10 (1.0 Pts) {-}
Compute the probability of obtaining 2, 3, 4, or 5 successes in 25 trials following a binomial distribution with probability of success being 0.25.

```{r}
round(
  sum(
    dbinom(2:5,25,.25)
    ),
  3)
```

### Question 11 (1.0 Pts) {-}
Consider participating in a lottery ten times. Each time the probability of winning a prize is 0.05. Compute the cumulative probability of winning 2 times or less rounded to three digits.

```{r}
round(
  pbinom(q = 2,
         size = 10,
         prob = .05),
  3
)

# this is a pretty high prob because there are large chances of winning 0 times
```

### Question 12 (1.0 Pts) {-}
Compute the value k such that the probability of obtaining k or fewer successes in 15 binomial samples with probability of success being 0.5 is 0.75.

```{r}
round(
  qbinom(.75,
         size = 15,
         prob = .5),
      3)
```

### Question 13 (1.0 Pts) {-}
Compute the value k such that the probability of obtaining k or fewer successes in 55 binomial samples with probability of success being 0.75 is 0.25.

```{r}
round(
  qbinom(.25,
         size = 55,
         prob = .75),
      3)
```

### Question 14 (1.0 Pts) {-}
What is the quantile q in a standard normal distribution, such that the probability of observing that value q or a value less than q is 0.25?

```{r}
round(
  qnorm(.25,
         mean = 0,
         sd = 1),
      3)
```

### Question 15 (1.0 Pts) {-}
Consider a normal distribution with mean 1 and standard deviation 1. Compute the boundary such that the area (the probability) to the left of it is 0.10.
```{r}
round(
  qnorm(p = .1,
         mean = 1,
         sd = 1),
      3)
```

### Question 16 (1.0 Pts) {-}
What is the probability of obtaining 6 successes in 18 trials following the binomial distribution with probability of success being 0.65?
```{r}
round(
  dbinom(x = 6,
         size = 18,
         p = .65),
      3)
```

### Question 17 (1.0 Pts) {-}
What is the probability of obtaining 17 successes in 18 trials following the binomial distribution with probability of success being 0.75?
```{r}
round(
  dbinom(x = 17,
         size = 18,
         p = .75),
      3)
```

### Question 18 (1.0 Pts) {-}
What is the probability of obtaining 6 successes in 60 trials following the binomial distribution with probability of success being 0.1?
```{r}
round(
  dbinom(x = 6,
         size = 60,
         p = .1),
      3)
```

### Question 19 (1.0 Pts) {-}

Given a normal distribution with mean 650 and standard deviation 125, there exist two quantiles, the lower quantile q1 and the upper quantile q2, that are equidistant from the mean, such that the area under the curve between q1 and q2 is 80%. Find q1 and q2.

```{r}
round(
  qnorm(p = c(.1,.9),
         mean = 650,
         sd = 125),
      3)
# alternative
q1 <- qnorm(p = .1, 650,125)
q2 <- qnorm(p = .1, 650,125, lower.tail=F)
```

# Week 2

## Question 1 (1.0 Pts) {-}

We ask a subject 10 yes/no questions, and the subject returns 0 correct answers. We assume a binomial likelihood function for this data. We also assume a Beta(1, 1) prior on the parameter θ, which represents the probability of success. Write down the posterior distribution of the θ parameter.

```{r}
# prior Beta(a,b)
a <- 1 
b <- 1 

# observed data
n <- 10 # trials
k <- 0 # successes

# posterior Beta equation (from section 2.7 in the book)
# posterior beta =
post_a <- a+k; post_a
post_b <- b + n - k; post_b
```

The answer is *Beta(*`r post_a`,`r post_b`*)*.

## Question 2 (1 Pt) {-}

Suppose you have n independent and identically distributed data points from the Geometric distribution, which has the likelihood function:

\begin{equation}
f(x|\theta)=\theta(1-\theta)^{\sum_{i=1}^n x_i}
\end{equation}

where the data points $x$ can have values 0, 1, 2, 3, … Let the prior on $\theta$ be $Beta (a, b)$, a beta distribution with parameters a, b. The posterior distribution is a beta distribution with parameters $a*$ and $b*$. Determine these parameters in terms of a, b and the sum of all x's.

Solution:

The prior is Beta(a,b).

The posterior likelihood is: $\theta(1 - \theta)^{\sum^{n}_{i=1^{x_{i}-1}}}$

The posterior up to proportionality is:

$[\theta^{a-1}(1 - \theta)^{b-1}][\theta(1 - \theta)^{\sum^{n}_{i=1^{x_{i}-1}}}]$

Re-arranging terms, where we add the exponents from either side (and where there is no exponent, we add '+1' because no exponent = 'to the power of 1'):

$\theta^{a + 1 - 1}(1 - \theta)^{b + \sum^{n}_{i=1^{x_{i}-1}}}$

If the posterior is a Beta distribution then its parameters are: $Beta(a + 1, b + \sum^{n}_{i=1^{x_{i}-1}})$. Thus:

$a* = a +1$

$b* = b + \sum^{n}_{i=1^{x_{i}-1}}$
